Week 2 coursework - Reproducible research
=============================================================

## read data from file
```{r}
data <- read.csv("activity.csv")
```

### Display the summary of the data
```{r}
summary(data)
```

### The mean steps across all days is
```{r}
mean(data$steps, na.rm=TRUE)
```

### The total number of steps across all days
```{r}
sum(data$steps, na.rm=TRUE)
```

## Calculate the total number of steps taken each day, removing NA values
```{r}
DSum <- tapply(data$steps ,data$date , sum, na.rm=TRUE)
```

## Create a histogram of the number of steps taken each day
```{r}
hist(DSum)
```

## Calculate the mean of total steps taken each day
```{r}
mean(DSum)
```

## Calculate the median of total steps taken each day
```{r}
median(DSum)
```

# What is the average daily activity pattern?

## Create a dataframe consisting of the mean values for each interval over all days
```{r}
S5Sum <- tapply(data$steps ,data$interval , mean, na.rm=TRUE)
S5Sum.T <- t(S5Sum)
stepsMean <- cbind(as.data.frame(S5Sum), colnames(S5Sum.T))
colnames(stepsMean) <- c("steps", "interval")
```

## Plot the Average steps taken across all days over each of the  intervals
```{r}
plot(stepsMean$steps,  ylab="Mean Steps", xlab="5 Min Interval",  type="l", main="Average steps taken across all days over each 5 min interval")
```

### Find the interval with the maximum number of average steps
```{r}
stepsMean[stepsMean[,1] == max(stepsMean$steps),]
```

# Imputing missing values

### Calculate the total number of missing values in the dataset
```{r}
nrow(data[is.na(data[,1]),])
```

### create a dataset of only missing values
```{r}
naData <-data[is.na(data[,1]),]
```

### create a merged data from the na-data and the mean data table using the mean values to impute the missing values
```{r}
mergedData = merge(naData,stepsMean, by.x="interval",by.y="interval",all=TRUE)
```

### create a cutdown version to match the original dataset
```{r}
newCreatedData <- data.frame(mergedData$steps.y,mergedData$date,mergedData$interval)
```

### set the column names of the new imputed values
```{r}
colnames(newCreatedData) <- c("steps", "date","interval")
```

### create a dataset that has no missing values from the original
```{r}
naData2 <-data[is.na(data[,1]) == FALSE,]
#head(naData2, 35)
```

### put the two datasets together
```{r}
newCreatedData2 <- rbind(newCreatedData, naData2)
```

## Summary of original dataset
```{r}
summary(data)
```

## Summary of imputed dataset
```{r}
summary(newCreatedData2)
```

## Calculate the total number of steps taken each day with imputed values  
```{r}
DSum2 <- tapply(newCreatedData2$steps ,newCreatedData2$date , sum, na.rm=TRUE)
```

## Histogram of the original dataset  
```{r}
hist(DSum)
```

## Histogram of the new dataset
```{r}
hist(DSum2)
```

## Mean of total steps each day - NEW dataset
```{r}
mean(DSum2)
```

## Median of total steps each day - NEW dataset
```{r}
median(DSum2)
```

## What is the impact of imupting vlaues RAW
```{r}
mean(DSum2)   # the avarage value in the new dataset
mean(DSum)    # the avarage value in the original dataset
median(DSum2) # the median value in the new dataset
median(DSum)  # the median value in the original dataset
```

### The mean increased by
```{r}
mean(DSum2) - mean(DSum)    

```

### The median increased by
```{r}
median(DSum2) - median(DSum)  
```


# Are there differences in activity patterns between weekdays and weekends?

### Use weekdays to add a new column to new dataset
```{r}
newCreatedData2 <- cbind(newCreatedData2, weekdays(as.Date(newCreatedData2$date)))
colnames(newCreatedData2) <- c("steps", "date","interval", "day")
```

### Create a new factor variable in the dataset with two levels - "weekday" and "weekend"
```{r}
newCreatedData2$weekend = chron::is.weekend(newCreatedData2$date)
```

# Weekend Plot
### create a plot of all average steps for the weekend
```{r}
plotdf <- newCreatedData2[newCreatedData2$weekend == TRUE, ]
```

### calculate the mean number of steps for each 5 min period - WEEKEND
```{r}
S5Sum <- tapply(plotdf$steps ,plotdf$interval , mean, na.rm=TRUE)
```

### time series plot, Average steps taken across all weekends over 288, 5 min interval periods
```{r}
plot(S5Sum,   ylab="Mean Steps", xlab="5 Min Interval Period",  type="l", main="Average steps taken across all weekends over 288, 5 min interval periods")
```

# Weekday Plot
### create a plot of all average steps for the weekdays
```{r}
plotdf <- newCreatedData2[newCreatedData2$weekend == FALSE, ] 
```

### calculate the mean number of steps for each 5 min period - WEEKDAY
```{r}
S5Sum <- tapply(plotdf$steps ,plotdf$interval , mean, na.rm=TRUE)
```

### time series plot, Average steps taken across all weekdays over 288, 5 min interval periods
```{r}
plot(S5Sum,  ylab="Mean Steps", xlab="5 Min Interval period",  type="l", main="Average steps taken across all weekdays over 288, 5 min interval periods")
```

